---
title: "groupproject"
author: "Yihang Zhang"
date: "4/24/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Trends for elapsed time

##Set up the matrix
```{r}
setwd("~/Stat480/RDataScience/Group")

library(biganalytics)
x <- read.big.matrix("groupprojectcl.csv", header = TRUE, 
                     backingfile = "gp.bin",
                     descriptorfile = "gp.desc",
                     type = "integer", extraCols = "Elapsed")
x1=x[x[,"Year"]==1998,]
x2=x[x[,"Year"]==2002,]

```

##Create the new variable for 1998
```{r}
for (i in 1:nrow(x1)) {
  x1[i,"Elapsed"]=x1[i,"ActualElapsedTime"]-x1[i,"CRSElapsedTime"]}

```

##Statistical summary for 1998
```{r}
over=x1[x1[,"Elapsed"]>0,]
nrow(over)
nrow(over)/nrow(x1)

intime=x1[x1[,"Elapsed"]<0,]
nrow(intime)
nrow(intime)/nrow(x1)

ontime=x1[x1[,"Elapsed"]==0,]
nrow(ontime)
nrow(ontime)/nrow(x1)

mean(x1[,"Elapsed"],na.rm=TRUE)
```

##Quantile plot for 1998
```{r}
myProbs <- seq(0.5, 0.99, by=0.01)
elapsedQuantiles=quantile(x1[, "Elapsed"], myProbs, 
                          na.rm = TRUE)
library(ggplot2)
library(reshape2) 
eq <- melt(elapsedQuantiles)
eq=cbind(percentile=rownames(eq), eq)
names(eq) <- c("percentile", "elapsed")
eq$percentile=as.double(eq$percentile)+50
qplot(percentile, elapsed, data = eq, color = percentile, geom = "line", title = "lol")

```

The plot demonstrates that there are 60% of the flights in 1998 spend no more than scheduled elapsed time to finish the journey. About 30% of flights spend under 10 minutes more than scheduled elapsed time, and there are 10% of flights spend above 10 minutes more than scheduled time, with the most reaching nearly 40 minutes.

##Create the new variable for 2002
```{r}
for (i in 1:nrow(x2)) {
  x2[i,"Elapsed"]=x2[i,"ActualElapsedTime"]-x2[i,"CRSElapsedTime"]}

```

##Statistical summary for 2002
```{r}
over2=x2[x2[,"Elapsed"]>0,]
nrow(over2)
nrow(over2)/nrow(x2)

intime=x2[x2[,"Elapsed"]<0,]
nrow(intime)
nrow(intime)/nrow(x2)

ontime=x2[x2[,"Elapsed"]==0,]
nrow(ontime)
nrow(ontime)/nrow(x2)

aveElap2002=mean(x2[,"Elapsed"],na.rm=TRUE)

```

##Quantile plot for 2002
```{r}
myProbs <- seq(0.5, 0.99, by=0.01)
elapsedQuantiles=quantile(x2[, "Elapsed"], myProbs, 
                          na.rm = TRUE)
library(ggplot2)
library(reshape2) 
eq <- melt(elapsedQuantiles)
eq=cbind(percentile=rownames(eq), eq)
names(eq) <- c("percentile", "elapsed")
eq$percentile=as.double(eq$percentile)+50
qplot(percentile, elapsed, data = eq, color = percentile, geom = "line")

```

The plot demonstrates that there are about 67% of the flights in 2002 spend no more than scheduled elapsed time to finish the journey. About 23% of flights spend under 10 minutes more than scheduled elapsed time, and there are 10% of flights spend above 10 minutes more than scheduled time, with the most also reaching nearly 40 minutes.

##Scatter plot for two year
```{r}

dayCount = integer(12)
for (i in 1:12) { 
  dayCount[i] <-  sum(over[,"Month"] == i, na.rm=TRUE)}
month=seq(1:12)
plot(month, dayCount, type = "o", col="blue", ylim=c(110000, 180000))

dayCount2 = integer(12)
for (i in 1:12) { 
  dayCount2[i] <-  sum(over2[,"Month"] == i, na.rm=TRUE)}
lines(dayCount2, type = "o", col="red")
title(main = "Days of over scheduled Elapsed time")

```

#Elapsed connecting with carriers

##Set up the data.frame
```{r}
groupprojectcsv=read.csv("groupprojectcl.csv")
groupprojectcl=groupprojectcsv[c("UniqueCarrier","ActualElapsedTime", "CRSElapsedTime")]
groupprojectcl$Elapsed=groupprojectcl$ActualElapsedTime-groupprojectcl$CRSElapsedTime
OvertimeEl=groupprojectcl[which(groupprojectcl$Elapsed>0),]

otcarrier=table(OvertimeEl$UniqueCarrier)
otcarrier=sort(otcarrier)
otcarrier=as.data.frame(otcarrier)
colnames(otcarrier) <- c("code", "freq")
```

##Merge the data by carrier code
```{r}
carriercsv=read.csv("carriers.csv")
OvertimeCarrier=merge(otcarrier, carriercsv, by.x="code", by.y="Code")
OvertimeCarrier[order(-OvertimeCarrier$freq),] 

```

#Elapsed connecting with origin airport

##Set up the data.frame
```{r}
airportcsv=read.csv("airports.csv")
groupprojectcl2=groupprojectcsv[c("Origin", "Dest", "ActualElapsedTime", "CRSElapsedTime")]
groupprojectcl2$Elapsed=groupprojectcl2$ActualElapsedTime-groupprojectcl2$CRSElapsedTime
OvertimeEl2=groupprojectcl2[which(groupprojectcl2$Elapsed>0),]

otorigin=table(OvertimeEl2$Origin)
otorigin=sort(otorigin)
otorigin=as.data.frame(otorigin)
colnames(otorigin) <- c("airport", "freq")
```

##Merge the data by airport code
```{r}
OvertimeOrigin=merge(otorigin, airportcsv, by.x="airport", by.y="iata")
OvertimeOrigin[order(-OvertimeOrigin$freq),] 
```

#Elapsed connectin with the destination airport

##Set up the data.frame
```{r}
otdest=table(OvertimeEl2$Dest)
otdest=sort(otdest)
otdest=as.data.frame(otdest)
colnames(otdest) <- c("airport", "freq")

```

##Merge the data by airport code
```{r}
OvertimeDest=merge(otdest, airportcsv, by.x="airport", by.y="iata")
OvertimeDest[order(-OvertimeDest$freq),] 

```

#Visualization: to compare the total flights number and over elapsed time flights number by carrier for the two-year data
```{r}
installIfNeeded(c("treemap"))
library(treemap)
carriermap=data.frame(table(groupprojectcl$UniqueCarrier))
colnames(carriermap) <- c("carrier", "freq")

AAcsv=groupprojectcl[which(groupprojectcl$UniqueCarrier=='AA'),]
mean(AAcsv$Elapsed, na.rm=TRUE)
AScsv=groupprojectcl[which(groupprojectcl$UniqueCarrier=='AS'),]
mean(AScsv$Elapsed, na.rm=TRUE)
COcsv=groupprojectcl[which(groupprojectcl$UniqueCarrier=='CO'),]
mean(COcsv$Elapsed, na.rm=TRUE)
DLcsv=groupprojectcl[which(groupprojectcl$UniqueCarrier=='DL'),]
mean(DLcsv$Elapsed, na.rm=TRUE)
HPcsv=groupprojectcl[which(groupprojectcl$UniqueCarrier=='HP'),]
mean(HPcsv$Elapsed, na.rm=TRUE)
MQcsv=groupprojectcl[which(groupprojectcl$UniqueCarrier=='MQ'),]
mean(MQcsv$Elapsed, na.rm=TRUE)
NWcsv=groupprojectcl[which(groupprojectcl$UniqueCarrier=='NW'),]
mean(NWcsv$Elapsed, na.rm=TRUE)
TWcsv=groupprojectcl[which(groupprojectcl$UniqueCarrier=='TW'),]
mean(TWcsv$Elapsed, na.rm=TRUE)
UAcsv=groupprojectcl[which(groupprojectcl$UniqueCarrier=='UA'),]
mean(UAcsv$Elapsed, na.rm=TRUE)
UScsv=groupprojectcl[which(groupprojectcl$UniqueCarrier=='US'),]
mean(UScsv$Elapsed, na.rm=TRUE)
WNcsv=groupprojectcl[which(groupprojectcl$UniqueCarrier=='WN'),]
mean(WNcsv$Elapsed, na.rm=TRUE)

carriermap$meanelapsed=c(-3.143742, 0.184173, -0.3499615, -0.2344786, -1.110613, -1.315059, -1.841802, -1.498755, -1.364657, -1.584349, -3.710075)

treemap(carriermap,
        index=c("carrier"),
        vSize="freq",
        vColor="meanelapsed",
        type="value")

```
Based on the plot, it is indicated that in terms of total flight volume, WN is the largeset, then DL, AA and UA. The flight number of these four carriers are accounted for more than half of the total flight number in 1998 and 2002. 
The other fact is that with the largest flight volumn, WN also has the least mean over elapsed time among all carriers, which is -3.5, meaning that WN's flights spend 3.5 minutes less than scheduled time on average. The second least one is AA, and then are the UW and TW. AS is the carrier seems to have the most serious over elapsed time problem of its flights.
